Fractional Types

### 2019-11-19 fracGC/p.tex
basic idea of pointed types in J.agda for presentation

\newunicodechar{𝕌}{$\mathbb{U}$}
\newunicodechar{𝔽}{$\mathbb{F}$}
\newunicodechar{𝕋}{$\mathbb{T}$}
\newunicodechar{𝟘}{$\mathbb{0}$}
\newunicodechar{𝟙}{$\mathbb{1}$}
\newunicodechar{𝟚}{$\mathbb{2}$}
\newunicodechar{⋆}{${}_*$}
\newunicodechar{ᵤ}{${}_u$}
\newunicodechar{⊚}{$\fatsemi$}

@@

\input{latex/PiFracDynCode.tex}

@@

\paragraph*{Outline.} The paper solves the ancilla problem in
reversible and quantum computation using a novel concept:
\emph{fractional types}. In the next section, we introduce the problem
of ancilla, motivate its importance, and explain the limitations of
current solutions. In Sec.~\ref{sec:pi}, we review previous work by
James et al.~\cite{rc2011,rc2012,James:2012:IE:2103656.2103667} that
introduces a reversible programming language built using isomorphism
of types. Although it is conceivable that the concept of fractional
types would apply to general-purpose languages (after some
adaptation), its natural technical definition is using symmetries
present in categorical model of type isomorphisms. We present such a
definition in Sec.~\ref{sec:dyn}: that definition captures the main
ideas of fractional types and partially solves a more general and more
expressive instance of the ancilla problem. The approach however
shares the limitation of current solutions in that it relies on
runtime checks for programmer's assertions that values have been
restored to their initial value before de-allocation. To remedy this
limitation, we refine the type system in Sec.~\ref{sec:dep} to keep
track of values in focus. Technically this requires the introduction
of pointed types on top of which are defined both a monad and co-monad
of singleton types. To justify the claim that fractional types solve
the management of ancilla-based scratch space, we present in
Sec.~\ref{sec:space} an operational semantics that exposes memory
management details using a realization of fractional types as
first-class dependently-typed garbage collectors. In Sec.~\ref{sec:cat} we
present a rich collection of examples illustrating the expressive
power of fractional types. We conclude with a discussion of related
work and a summary of our results.

@@

\begin{center}
\begin{tikzpicture}[scale=0.4,every node/.style={scale=0.5}]
  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
  \draw (-2,-2) ellipse (0.9cm and 1cm);
  \draw[fill] (-2,-1.5) circle [radius=0.025];
  \node[left] at (-2.1,-1.5) {$A$};
  \draw[fill] (-2,-2.5) circle [radius=0.025];
  \node[left] at (-2.1,-2.5) {$B$};

@@

%%%
\subsection{A Small Example}

some that ``wants'' ancilla; perhaps fig2b from Dyn. if we want to use
gates with small number of inputs; need to break cccnot to smaller gates

@@

This scoped model is therefore quite a pragmatic choice. It is however
limited, and to understand its limitations more vividly, we propose
the following analogy: allocating an ancilla by creating a new wire in
the circuit is like borrowing some money from a global external
entity; the computation has access to a new resource
temporarily. De-allocating the ancilla is like returning the borrowed
money to the global entity; the computation no longer has access to
the additional resource. It would however be unreasonably restrictive
to insist that if the person (function) borrowing the money must be
the same person (function) returning it.
%% , i.e., preventing the following computation:
%% \begin{verbatim}
%%   |-----\  /-----|
%%          \/
%%         / \
%%   |----/   \-----|
%% \end{verbatim}
Indeed, as far as reversible or quantum computation is concerned, the
only important invariant is that information is conserved, i.e., that
money is conserved. The identities of bits or qubits are not
observable as they are all interchangeable in the same way that
particular bills with different serial numbers are interchangeable in
financial transactions. Thus the only invariant is that the net flow
of money between the computation and the global entity is zero. This
observation allows us to go even further than just switching the
identities of borrowers. It is even possible for one person to borrow
\$10, and have three different persons collectively collaborate to
pay back the debt with one person paying \$5, another \$2, and a third
\$3.

Computationally, this extra generality is not a gratuitous concern:
since scope is fundamentally a static property of programs, it does
not allow the flexibility of heap allocation in which the lifetime of
resources is dynamically determined. To summarize, since the
reversible-quantum computational framework guarantees that information
is preserved, it also permits fascinating mechanisms in which
computations can be sliced and diced, decomposed and recomposed, run
forwards and backwards, in arbitrary ways as long as the net balance
is 0.

>>>>

\newunicodechar{𝕌}{$\mathbb{U}$}
\newunicodechar{𝔹}{$\mathbb{B}$}
\newunicodechar{𝔽}{$\mathbb{F}$}
\newunicodechar{𝕋}{$\mathbb{T}$}
\newunicodechar{𝟘}{$\mathbb{0}$}
\newunicodechar{𝟙}{$\mathbb{1}$}
\newunicodechar{𝟚}{$\mathbb{2}$}
\newunicodechar{⋆}{${}_*$}
\newunicodechar{ᵤ}{${}_u$}
\newunicodechar{⊚}{$\fatsemi$}
\newunicodechar{⟷}{$\leftrightarrow$}

@@

\input{latex/PiFracDynCode.tex}
\input{latex/BooleanCircuitsCode.tex}

@@

\paragraph*{Outline.} The paper solves the ancilla problem in
reversible and quantum computation using a novel concept:
\emph{fractional types}. In the next section, we introduce the problem
of ancilla, motivate its importance, and explain the limitations of
current solutions. In Sec.~\ref{sec:pi}, we review existing work by
James et al.~\cite{rc2011,rc2012,James:2012:IE:2103656.2103667} that
introduces a reversible programming language built using isomorphism
of types. Although it is conceivable that the concept of fractional
types would apply to general-purpose languages (after some
adaptation), its natural technical definition is using symmetries
present in categorical model of type isomorphisms. We present such a
definition in Sec.~\ref{sec:dyn}: that definition captures the main
ideas of fractional types and partially solves a more general and more
expressive instance of the ancilla problem. The approach however
shares the limitation of current solutions in that it relies on
runtime checks for programmer's assertions that values have been
restored to their initial value before de-allocation. To remedy this
limitation, we refine the type system in Sec.~\ref{sec:dep} to keep
track of values in focus. Technically this requires the introduction
of pointed types on top of which are defined both a monad and co-monad
of singleton types. To justify the claim that fractional types solve
the management of ancilla-based scratch space, we present in
Sec.~\ref{sec:space} an operational semantics that exposes memory
management details using a realization of fractional types as
first-class dependently-typed garbage collectors. In
Sec.~\ref{sec:cat} we present a rich collection of examples
illustrating the expressive power of fractional types. We conclude
with a discussion of related work and a summary of our results.

@@

\begin{center}
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.6}]
  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
  \draw (-2,-2) ellipse (0.9cm and 1cm);
  \draw[fill] (-2,-1.5) circle [radius=0.025];
  \node[left] at (-2.1,-1.5) {$A$};
  \draw[fill] (-2,-2.5) circle [radius=0.025];
  \node[left] at (-2.1,-2.5) {$B$};

@@

%%%
\subsection{Small Examples}

The code below defines the types corresponding to bits (booleans),
two-bits, three-bits, and four-bits. It then defines an operator that
builds a controlled version of a given combinator $c$. This controlled
version takes an additional ``control'' bit and only applies $c$ if
the control bit is true. The code then iterates the control operation
several times starting starting from boolean negation.

\Bexamples{}

By construction, the \AgdaFunction{TOFFOLI} function takes three bits,
and negates the third if the first two are both true. As
Toffoli~\cite{Toffoli:1980} demonstrated, this function is universal
for classical reversible circuits assuming it is possible to allocate
and de-allocate ancilla bits. Indeed without using ancilla bits, it is
not possible to use compositions of \AgdaFunction{TOFFOLI} functions
to implement more general versions that are controlled by more than
two bits (for example, \AgdaFunction{CTOFFOLI}). This will however be
possible in the next section once we introduce ancilla bits.

@@

This scoped model is therefore quite a pragmatic choice. It is however
limited, and to understand its limitations more vividly, we propose
the following analogy: allocating an ancilla by creating a new wire in
the circuit is like borrowing some money from a global external
entity (the memory manager); the computation has access to a new resource
temporarily. De-allocating the ancilla is like returning the borrowed
money to the global entity; the computation no longer has access to
the additional resource. It would however be unreasonably restrictive
to insist that the person (function) borrowing the money must be
the same person (function) returning it.
%% , i.e., preventing the following computation:
%% \begin{verbatim}
%%   |-----\  /-----|
%%          \/
%%         / \
%%   |----/   \-----|
%% \end{verbatim}
Indeed, as far as reversible or quantum computation is concerned, the
only important invariant is that information is conserved, i.e., that
money is conserved. The identities of bits or qubits are not
observable as they are all interchangeable in the same way that
particular bills with different serial numbers are interchangeable in
financial transactions. Thus the only invariant is that the net flow
of money between the computation and the global entity is zero. This
observation allows us to go even further than just switching the
identities of borrowers. It is even possible for one person to borrow
\$10, and have three different persons collectively collaborate to
pay back the debt with one person paying \$5, another \$2, and a third
\$3.

Computationally, this extra generality is not a gratuitous concern:
since scope is fundamentally a static property of programs, it does
not allow the flexibility of heap allocation in which the lifetime of
resources is dynamically determined. To summarize, since the
reversible-quantum computational framework guarantees that information
is preserved, it also permits fascinating mechanisms in which
allocations and de-allocations can be sliced and diced, decomposed and
recomposed, run forwards and backwards, in arbitrary ways as long as
the net balance is 0.
